### 2010-02-24 trunk/deliverables/D2_2_Ontology/classes.tex
making more improvements, added piping example figure
git-svn-id: https://svn.forge.morfeo-project.org/fast-fp7project@597 d8b0c59b-6544-0410-94d9-83a5658a058e

\textcolor{white}{\textbf{in\_domain\_of}} & \htmlref{\texttt{fgo:contains}}{subs:contains}, \htmlref{\texttt{fgo:hasIcon}}{subs:hasIcon}, \htmlref{\texttt{fgo:hasScreenshot}}{subs:hasScreenshot} \\ \hline
\textcolor{white}{\textbf{in\_range\_of}} & \htmlref{\texttt{fgo:contains}}{subs:contains} \\ \hline

>>>>

\textcolor{white}{\textbf{in\_domain\_of}} & \htmlref{\texttt{fgo:contains}}{subs:contains}, \htmlref{\texttt{fgo:hasCopy}}{subs:hasCopy}, \htmlref{\texttt{fgo:hasIcon}}{subs:hasIcon}, \htmlref{\texttt{fgo:hasScreenshot}}{subs:hasScreenshot}, \htmlref{\texttt{fgo:hasTemplate}}{subs:hasTemplate} \\ \hline
\textcolor{white}{\textbf{in\_range\_of}} & \htmlref{\texttt{fgo:contains}}{subs:contains}, \htmlref{\texttt{fgo:hasCopy}}{subs:hasCopy}, \htmlref{\texttt{fgo:hasTemplate}}{subs:hasTemplate} \\ \hline

### 2010-02-24 trunk/deliverables/D2_2_Ontology/fast_d2_2_2_ontology.tex
making more improvements, added piping example figure
git-svn-id: https://svn.forge.morfeo-project.org/fast-fp7project@597 d8b0c59b-6544-0410-94d9-83a5658a058e

% This is useful for switching in-line comments on and off!
\usepackage{ifthen}
\newboolean{showcomments}
\setboolean{showcomments}{false}
\ifthenelse{\boolean{showcomments}}
{
	\definecolor{Red}{rgb}{1,0,0}
	\newcommand{\note}[1]{\begin{scriptsize}{{\textsf{\textbf{{\textcolor{Red}{[[#1]]}}}}}}\end{scriptsize}}
	\definecolor{Orange}{rgb}{1,0.5,0}
	\newcommand{\todo}[1]{\textsf{\textbf{\textcolor{Orange}{[[TODO: #1]]}}}}
    \definecolor{Green}{rgb}{0,1,0}
    \newcommand{\commentreply}[1]{\begin{scriptsize}{\textsf{\textbf{\textcolor{Green}{[[#1]]}}}}\end{scriptsize}}
} {
	\newcommand{\note}[1]{\textbf{}}
	\newcommand{\todo}[1]{}
    \newcommand{\commentreply}[1]{}
}

@@

\begin{itemize}
	\item \textbf{Abstract Class Level}: At this level, resources are defined on a categorical level, differentiating between the various kinds of things that are relevant in FAST, as fundamental classes such as screens, forms or resources (see Sect.~\ref{sub:conceptualisation}). These are implemented technically as classes in OWL.
	\item \textbf{Prototypical Component Definition}: When a user of the FAST GVS is building a new screenflow or screen, they do so by combining existing building blocks and configuring them. These building blocks are not the fundamental classes of FAST, but instead they are specific instances of those classes. In other words, users do not combine the idea of a screen with the idea of a service, but instead they combine the specific ``Log In Screen'' with the ``Amazon Web Service''. Technically, these concrete building blocks are implemented as instances of the fundamental classes, specified by asserting particular settings such as label or icons, pre- and post-conditions, etc.
	\item \textbf{Instantiation within a Screenflow at Design-time}: Once a user selects a specific building block in the GVS and adds it to their new screen or screenflow project, they are in principle again instantiating from a conceptual to a concrete level: from the ``Log In Screen'' as such to a particular log in screen as used in that particular screenflow. Also in this stage, resources are implemented technically as instances.
	\item \textbf{Runtime}: At the final level of instantiation, the screenflow has been packaged as a gadget, deployed to a gadget platform and is now being used by an end-user. At this level, all gadget components exist only in the form of Web standards compliant code (HTML, JavaScript, CSS). Any metadata still necessary or definitions for screen pre- and post-conditions are for performance now defined in the JavaScript-native data format JSON, rather than in RDF. Therefore, at this level, the resources that make up the gadget are out of the scope of this deliverable.
\end{itemize}

@@

Neither OWL nor RDFS semantics offer functionality to represent a relationship such as the copy of a resource formally. For this reason, we introduce the concept of \emph{templates for RDF instances} in FAST. One resource (e.g., $LogInScreen$) will play the role of a \emph{template}, while the other resource (e.g., $LogInScreen_x$) will be called a \emph{copy}. Conceptually, copying a resource means defining a named graph that contains all statements about this resource considered to be relevant, and creating a new graph with near-identical statements\footnote{The graphs are not completely identical: the identifiers for the resource and potentially other, contained resources will have changed.} \todo{explain this better}. In practice, this procedure can be implemented using a SPARQL CONSTRUCT query, where the WHERE clause matches the graph pattern of the template, and the CONSTRUCT query will build the copy from it. The corresponding data on all three levels is shown in List.~\ref{list:templating}, using TriG syntax~\cite{bizer2004trig} (an extension of Turtle syntax for named graphs).

\singlespacing
\lstset{
	caption={Templating example --- different levels of representation of a FAST screen}, 
	label=list:templating,
	language=turtle
}
\begin{figure}[ht]
	\lstinputlisting{exampleCode/templating_example.ttl}
\end{figure}
\doublespacing

@@

Once a user has been defined in this way, they can e.g. be used to specify the creators of individual screen and screenflow resources, as shown in List.~\ref{list:creator}. This kind of creation metadata is being added to other metadata as shown in Sect.~\ref{sub:basic_annotation}.

@@

@@

\singlespacing
\begin{itemize}
	\item \textbf{reuse external ontologies} to facilitate integration and compatibility,
	\item \textbf{avoid over-engineering} and keep the number of ontology terms down to reduce the cognitive load on implementers,
	\item \textbf{be a good Web citizen} by following the principles of linked data,
	\item \textbf{be well documented} to make it easy for implementers to understand and use the ontology correctly and
	\item \textbf{provide tool support} to ease the work-load involved in generating instance data.
\end{itemize}
\doublespacing

@@

\subsubsection{Be Well Documented} % (fold)
\label{ssub:be_well_documented}

@@

In order to put the design of the ontology on a stable footing, we chose to adopt a tried and tested design methodology, rather than using a purely ad-hoc approach. We decided to use the \emph{Methontology} methodology, which was first introduced in \cite{gomez_perez1996conceptualise_ontologies} and \cite{fernandez1997methontology}. Methontology provides a good balance of formalisation and streamlining of the design process on the one hand, and on the other hand a looseness in its requirements on how strictly one needs to follow the individual phases and steps. The authors specifically advocate an ``evolving prototype life cycle'', which allows the ontology to ``grow depending on its needs'' \cite{fernandez1997methontology}.

>>>>

% This is useful for switching in-line comments on and off!
\usepackage{ifthen}
\newboolean{showcomments}
\setboolean{showcomments}{true}
\ifthenelse{\boolean{showcomments}}
{
	\definecolor{Red}{rgb}{1,0,0}
	\newcommand{\note}[1]{\begin{scriptsize}{{\textsf{\textbf{{\textcolor{Red}{[[#1]]}}}}}}\end{scriptsize}}
	\definecolor{Orange}{rgb}{1,0.5,0}
	\newcommand{\todo}[1]{\textsf{\textbf{\textcolor{Orange}{[[TODO: #1]]}}}}
    \definecolor{Green}{rgb}{0,1,0}
    \newcommand{\commentreply}[1]{\begin{scriptsize}{\textsf{\textbf{\textcolor{Green}{[[#1]]}}}}\end{scriptsize}}
} {
	\newcommand{\note}[1]{\textbf{}}
	\newcommand{\todo}[1]{}
    \newcommand{\commentreply}[1]{}
}

@@

\begin{itemize}
	\item \textbf{Abstract Class Level}: At this level, resources are defined on a categorical level, differentiating between the various kinds of things that are relevant in FAST, as fundamental classes such as screens, forms or resources (see Sect.~\ref{sub:conceptualisation}). These are implemented technically as classes in OWL.
	\item \textbf{Prototypical Component Definition}: When a user of the FAST GVS is building a new screenflow or screen, they do so by combining existing building blocks and configuring them. These building blocks are not the fundamental classes of FAST, but instead they are specific instances of those classes. In other words, users do not combine the idea of a screen with the idea of a service, but instead they combine the specific ``Log In Screen'' with the ``Amazon Web Service''. Technically, these concrete building blocks are implemented as instances of the fundamental classes, specified by asserting particular settings such as label or icons, pre- and post-conditions, etc.
	\item \textbf{Instantiation within a Screenflow at Design-time}: Once a user selects a specific building block in the GVS and adds it to their new screen or screenflow project, they are in principle again instantiating from a conceptual to a concrete level: from the ``Log In Screen'' as such to a particular log in screen as used in that particular screenflow. Also in this stage, resources are implemented technically as instances.
	\item \textbf{Runtime}: At the final level of instantiation, the screenflow has been packaged as a gadget, deployed to a gadget platform and is now being used by an end-user. At this level, all gadget components exist only in the form of Web standards compliant code (HTML, JavaScript, CSS). For performance reasons, the FAST catalogue transforms any metadata still necessary, as well as definitions for screen pre- and post-conditions, from their original RDF representations into the JavaScript Object Notation (JSON)\footnote{\url{http://www.json.org/}}. At this level, the resources that make up the gadget have moved out of the scope of the FAST platform and therefore out of the scope of this deliverable.
\end{itemize}

@@

Neither OWL nor RDFS semantics offer functionality to represent a relationship such as the copy of a resource formally. For this reason, we introduce the concept of \emph{templates for RDF instances} in FAST. One resource (e.g., $LogInScreen$) will play the role of a \emph{template}, while the other resource (e.g., $LogInScreen_x$) will be called a \emph{copy}. Conceptually, copying a resource involves the following steps:
\begin{inparaenum}[(i)]
	\item Defining a named graph $G_1$ that contains all statements about this resource considered to be relevant.
	\item Creating a second graph $G_2$ with copies of all statements in $G_1$.
	\item During this copying process, the identifiers for the main resource and certain other resources in the graph will be changed from $I_{template}$ to $I_{copy}$.
\end{inparaenum}	
In practice, this procedure can be implemented using a SPARQL CONSTRUCT query, where the WHERE clause defines the graph pattern of the template, and the CONSTRUCT query will build the copy from it. The corresponding data on all three levels is shown in List.~\ref{list:templating}, using TriG syntax~\cite{bizer2004trig} (an extension of Turtle syntax for named graphs).

\singlespacing
\lstset{
	caption={Templating example --- different levels of representation of a FAST screen (TriG)}, 
	label=list:templating,
	language=turtle
}
\begin{figure}[ht]
	\lstinputlisting{exampleCode/templating_example.ttl}
\end{figure}
\doublespacing

Figure \ref{fig:templating_example_graph} illustrates the same example graphically, using different colours to represent the two different graphs: the red graph in the background represents the original template resource, whereas the green graph in the front in represents the resource copy as it would be used in a concrete screen flow.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=.75\linewidth]{images/templating_example.pdf}
    \caption{Templating example --- different levels of representation of a FAST screen (Diagramme)}
    \label{fig:templating_example_graph}
  \end{center}
\end{figure}

@@

Once a user has been defined in this way, they can for example be used to specify the creators of individual screen and screenflow resources, as shown in List.~\ref{list:creator}. This kind of creation metadata is being added to other metadata as shown in Sect.~\ref{sub:basic_annotation}.

@@

\subsection{Piping within Screens} % (fold)
\label{sub:piping_within_screens}

\todo{needs text and code}

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=\linewidth]{images/declarativeDefinitionExample.pdf}
    \caption{Explicit piping within a screen from a post- to a pre-condition}
    \label{fig:piping}
  \end{center}
\end{figure}

% subsection piping_within_screens (end)

@@

\singlespacing
\begin{itemize}
	\item \textbf{reuse external ontologies} to facilitate integration and compatibility,
	\item \textbf{avoid over-engineering} and keep the number of ontology terms down to reduce the cognitive load on implementers,
	\item \textbf{be a good Web citizen} by following the principles of linked data,
	\item \textbf{document well} to make it easy for implementers to understand and use the ontology correctly and
	\item \textbf{provide tool support} to ease the work-load involved in generating instance data.
\end{itemize}
\doublespacing

@@

\subsubsection{Document Well} % (fold)
\label{ssub:be_well_documented}

@@

In order to put the design of the ontology on a stable footing, we chose to adopt a tried and tested design methodology, rather than using a purely ad-hoc approach. We decided to use the \emph{Methontology} methodology, which was first introduced in \cite{gomez_perez1996conceptualise_ontologies} and \cite{fernandez1997methontology}. Methontology provides a good balance of formalisation and streamlining of the design process on the one hand, while on the other hand its requirements on how strictly one needs to follow the individual phases and steps are intentionally left relatively loose.
The authors specifically advocate an ``evolving prototype life cycle'', which allows the ontology to ``grow depending on its needs'' \cite{fernandez1997methontology}.

### 2010-02-24 trunk/deliverables/D2_2_Ontology/properties.tex
making more improvements, added piping example figure
git-svn-id: https://svn.forge.morfeo-project.org/fast-fp7project@597 d8b0c59b-6544-0410-94d9-83a5658a058e

@@

>>>>

\subsubsection*{Property: \texttt{fgo:hasCopy}}
\label{subs:hasCopy}
\textcolor{white}{\textbf{label}} & has copy \\ \hline
\textcolor{white}{\textbf{description}} & Links the template of a resource as used in the palette or catalogue of available resources to its copy, as used in a particular screen flow. \\ \hline

@@

\subsubsection*{Property: \texttt{fgo:hasTemplate}}
\label{subs:hasTemplate}
\textcolor{white}{\textbf{label}} & has template \\ \hline
\textcolor{white}{\textbf{description}} & Links the copy of a resource as used in a particular screen flow to its template, as used in the palette or catalogue of available resources. \\ \hline
